---
title: "Modify phylogenetic trees"
author: "Lisa"
date: "10/11/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
packages <- c("ape","magrittr","TreeTools","tibble")
for (package in packages){
  if (!require(package, character.only = TRUE)) install.packages(package)
  require(package, character.only = TRUE)
}
```

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
packages <- c("ggtree")
for (package in packages){
  if (!require(package, character.only = TRUE)) BiocManager::install(package)
  require(package, character.only = TRUE)
}
```

### Import phylogenetic trees
```{r}
nj_tree <- ape::read.tree("../results/panaroo_output_all/nj_snp_tree.newick")
ml_tree <- ape::read.tree("../results/panaroo_output_all/RAxML_bipartitions.bootstrap_tree")
```

Let's take a look
```{r}
ggtree(nj_tree)
ggtree(ml_tree)
```
We already see the trees produced by the different methods are quite different.

### Modify trees
There is one sample that has an extremely long edge length, it seems to be a species other than E. coli entirely.
We can exclude it.
```{r}
nj_tree_dropped <- ape::drop.tip(nj_tree,"ECO-JSC-RGN-103823")
ml_tree_dropped <- ape::drop.tip(ml_tree,"ECO-JSC-RGN-103823")
```

The trees are unrooted. We can use the divergent sample as the root before excluding it.
```{r}
nj_tree_rooted <- root(nj_tree,"ECO-JSC-RGN-103823") %>% ape::drop.tip("ECO-JSC-RGN-103823")
ml_tree_rooted <- root(ml_tree,"ECO-JSC-RGN-103823") %>% ape::drop.tip("ECO-JSC-RGN-103823")
```
Let's look at the rooted trees
```{r}
ggtree(nj_tree_rooted,layout = "circular")
ggtree(ml_tree_rooted, layout = "circular")
```

I want to see the bootstrap support values
Code adjusted from: https://gist.github.com/joelnitta/97a03ab1f32af99cc0d80ade4e4a444b

Make a tibble with the nodes and bootstrap values:
```{r}
bs_tibble <- tibble(
  node=1:Nnode(ml_tree_rooted) + Ntip(ml_tree_rooted),
  bootstrap = ml_tree_rooted$node.label
)
```

Plot them in the tree:
```{r}
ggtree(ml_tree_rooted) %<+% bs_tibble +
  geom_text(aes(label=bootstrap), hjust=-.25, size = 3)
```
The bootstrap values of the first two splits are 0, probably because of rooting the tree ourselves.
Compared the original tree:
```{r}
ggtree(ml_tree) %<+% bs_tibble +
  geom_text(aes(label=bootstrap), hjust=-.25, size = 3)
```
The first 0 is due to rooting, the second 0 was already there in the original tree. It is good to know that there is absolutely no basis for this split.

Save trees
```{r}
write.tree(nj_tree_rooted,"../results/final_nj_tree.newick")
write.tree(ml_tree_rooted,"../results/final_ml_tree.newick")
```

### Make ST131 tree
Cut the tree
```{r}
ST131_strains <- read.delim("../results/all_ST131_rgnosis.txt",header = FALSE)
ST131_labels <- ST131_strains$V1
ml_ST131 <- ape::keep.tip(ml_tree_rooted,ST131_labels)
ggtree(ml_ST131, layout = "circular")
```
Write out
```{r}
write.tree(ml_ST131,"../results/ST131_ml_tree.newick")
```

